---
title: "MPP Capstone Data Analysis"
author: "Kevin Sullivan"
date: '2025-02-26'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(readr)
library(dplyr)
library(skimr)
library(foreign)
library(ggplot2)
library(GGally)
library(naniar)
library(kableExtra)
library(summarytools)
library(stargazer)
library(gtsummary)
library(tidyr)
```

```{r}
# Read in MEPS Data Previous 2017
MEPS_2005 = read.xport("h97.ssp")
MEPS_2006 = read.xport("h105.ssp")
MEPS_2007 = read.xport("h113.ssp")
MEPS_2008 = read.xport("h121.ssp")
MEPS_2009 = read.xport("h129.ssp")
MEPS_2010 = read.xport("h138.ssp")
MEPS_2011 = read.xport("h147.ssp")
MEPS_2012 = read.xport("h155.ssp")
MEPS_2013 = read.xport("h163.ssp")
MEPS_2014 = read.xport("h171.ssp")
MEPS_2015 = read.xport("h181.ssp")
MEPS_2016 = read.xport("h192.ssp")
```

```{r}
# Read in MEPS Data After 2017
MEPS_2017 = read.csv("h201.csv")
MEPS_2018 = read.csv("h209.csv")
MEPS_2019 = read.csv("h216.csv")
MEPS_2020 = read.csv("h224.csv")
MEPS_2021 = read.csv("h233.csv")
MEPS_2022 = read.csv("MEPS_2022.csv")
```

**Preprocessing for diff-in-diff (Data Cleaning, Merging, Variable Creation)**


 Variables of Interest 

 Outcome Variables: OBDRV = # of doctor visits, HAVEUS = indicator if respondent has a usual source of care, ADRTCR = made appointment for routine care
 
 Treatment and Control Groups: MCDEV (1 if has medicaid 0 else), 1 = treatment 0 = control
 
 Time = 2014 (Medicaid expansion)
 
 Pre = Years before 2014
 
 Post = Years after 2014
 
 so ADRTCR42, HAVEUS42 is consistent across all years

 MCDEV, OBDRV changes by year

```{r}
# Write Function to Clean Data, Select Desired Variables, and Create Necessary diff-in-diff variables
meps_list = list(
  MEPS_2005, MEPS_2006, MEPS_2007, MEPS_2008, MEPS_2009, MEPS_2010, MEPS_2011, MEPS_2012,
  MEPS_2013, MEPS_2014, MEPS_2015, MEPS_2016, MEPS_2017, MEPS_2018, MEPS_2019, MEPS_2020,
  MEPS_2021, MEPS_2022
)
clean_meps = function(df, year) {
  
  df = df %>%
    select(
      DUPERSID, 
      starts_with("OBDRV"),  # Doctor visits
      starts_with("HAVEUS"), # Usual source of care
      starts_with("MCDEV"),  # Medicaid enrollment
      starts_with("ADRTC"),  # Appointment for Routine Care
      starts_with("AGE"),    # Age
      starts_with("TTLP"),   # Income
      starts_with("RACE"),   # Race/Ethnicity
      starts_with("SEX"),    # Gender
      starts_with("EDUC"),   # Education
      starts_with("EMPST"), # Employment status
      starts_with("REGION"),  # Geographic region
      starts_with("ADBMI")  # BMI
    )
  
  # Remove year suffixes 
  colnames(df) = gsub("([A-Za-z]+)[0-9]+", "\\1", colnames(df))

  df = df[, !duplicated(colnames(df))]

  # Add year variable
  df = df %>%
    mutate(YEAR = year)

  # Define Treatment Group based on Medicaid coverage
  if ("MCDEV" %in% colnames(df)) {
    df = df %>%
      mutate(Treatment_Group = ifelse(MCDEV == 1, 1, 0))  
  }

  # Create Pre/Post Expansion Indicator
  df = df %>%
    mutate(Pre_Post = ifelse(YEAR < 2014, "Pre", "Post"))  

  # Ensure DUPERSID is a character
  df = df %>%
    mutate(DUPERSID = as.character(DUPERSID))
  
  return(df)
}

meps_cleaned_list = lapply(2005:2022, function(year) clean_meps(meps_list[[year - 2004]], year))

meps_data = bind_rows(meps_cleaned_list)

meps_data = meps_data %>%
  mutate(
    HAVEUS = ifelse(HAVEUS == 1, 1, 0),  
    ADRTCR = ifelse(ADRTCR == 1, 1, 0)     
  )

meps_data = meps_data %>%
  select(c("DUPERSID", "OBDRV", "HAVEUS", "MCDEV", "ADRTCR", "AGEX", "TTLPX", "RACEVX", "RACEBX", "RACEAX", "RACEWX","SEX", "EDUCYR", "EMPST", "REGION", "YEAR", "Treatment_Group", "Pre_Post")) %>%
  rename(Income = TTLPX) %>% 
  mutate(
    RACEVX = case_when(
      RACEVX == 1 ~ "White",
      RACEVX == 2 ~ "Black",
      RACEVX == 4 ~ "Asian",
      RACEVX == 3 | RACEVX == 5 | RACEVX == 6 ~ "Other",
      is.na(RACEVX) & RACEBX == 1 ~ "Black",
      is.na(RACEVX) & RACEWX == 1 ~ "White",
      is.na(RACEVX) & RACEAX == 1 ~ "Asian",
      TRUE ~ "Unknown"  
    ),
    REGION = case_when(
      REGION == 1 ~ "Northeast",
      REGION == 2 ~ "Midwest",
      REGION == 3 ~ "South",
      REGION == 4 ~ "West",
      TRUE ~ "Unknown"   
    )
  )

meps_data = meps_data %>%
  select("DUPERSID", "OBDRV", "HAVEUS", "MCDEV", "ADRTCR", "AGEX", "Income", "RACEVX", "SEX", "EDUCYR", "EMPST", "REGION", "YEAR", "Treatment_Group", "Pre_Post") %>%
  rename(ID = DUPERSID, Number_Doctor_Visits = OBDRV, Have_Usual_Source_of_Care = HAVEUS, Made_Appointment_for_Routine_Care = ADRTCR, Medicaid = MCDEV, Age = AGEX, Race = RACEVX, Sex = SEX, Years_of_Education = EDUCYR, Employment_Status = EMPST, Region = REGION, Year = YEAR) %>%
  mutate( 
    Medicaid = ifelse(Medicaid == 2, 0, Medicaid),
    Sex = case_when(Sex == 1 ~ "Male",
                    Sex == 2 ~ "Female")
    )

head(meps_data)
```
**EDA**

```{r}
# Percent Missing Data
meps_missing_by_var = colMeans(is.na(meps_data)) * 100

meps_missing_by_var

```


```{r}
# Numeric Summary Stats Table

stargazer(meps_data, type = "html", title = "Table 1: Numeric Descriptive Statistics", digits = 1, out = "table1.html", column.sep.width = "5pt")

cat("<style>table { border-spacing: 10px; }</style>", file = "table1.html", append = TRUE)
```

```{r}
# Categorical Summary Stats Table

categorical_vars = meps_data %>% select(where(~is.factor(.) | is.character(.)))

categorical_vars = categorical_vars %>% mutate(across(everything(), as.factor))

dfSummary(categorical_vars) %>%
  print(method = "browser")

```


**Graph Pre and Post Trends for Each Outcome Variable**

```{r}
trend_data = meps_data %>%
  group_by(Year, Pre_Post, Treatment_Group) %>%
  summarise(
    Mean_doctor_visits = mean(Number_Doctor_Visits, na.rm = TRUE),
    Mean_haveus = mean(Have_Usual_Source_of_Care, na.rm = TRUE),
    Mean_appointment = mean(Made_Appointment_for_Routine_Care, na.rm = TRUE)
  ) %>%
  ungroup()

head(trend_data)
```
```{r}
# Graph for Doctor Visits

ggplot(trend_data, aes(x = Year, y = Mean_doctor_visits, color = as.factor(Treatment_Group), linetype = Pre_Post)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Pre and Post Trends for Number of Doctor Visits",
    x = "Year",
    y = "Mean Number of Doctor Visits",
    color = "Treatment Group",
    linetype = "Pre/Post"
  ) + 
  scale_color_manual(
    values = c("mediumblue", "deepskyblue1"),
    labels = c("Non-Medicaid", "Medicaid")  # Labeling groups manually
  ) +
  geom_vline(xintercept = 2014, linetype = "dotted", color = "black", size = 0.5) +
  geom_text(aes(x = 2014, y = -Inf, label = "Medicaid Expansion"), 
            angle = 0, hjust = 0.46, vjust = -0.5, size = 3, color = "firebrick2") +
  theme_minimal()


```

```{r}
# Graph for HAVEUS

ggplot(trend_data, aes(x = Year, y = Mean_haveus, color = as.factor(Treatment_Group), linetype = Pre_Post)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Pre and Post Trends for Respondent Having a Primary Care Provider",
    x = "Year",
    y = "Response to Having a Primary Care Provider (1 = Yes, 0 = No)",
    color = "Treatment Group",
    linetype = "Pre/Post"
  ) + 
  scale_color_manual(values = c("mediumblue", "deepskyblue1")) +
  scale_color_manual(
    values = c("mediumblue", "deepskyblue1"),
    labels = c("Non-Medicaid", "Medicaid")  # Labeling groups manually
  ) +
  geom_vline(xintercept = 2014, linetype = "dotted", color = "black", size = 0.5) +
  geom_text(aes(x = 2014, y = -Inf, label = "Medicaid Expansion"), 
            angle = 0, hjust = 0.46, vjust = -0.5, size = 3, color = "firebrick2") +
  theme_minimal()

```

```{r}
# Graph for ADTCR

ggplot(trend_data, aes(x = Year, y = Mean_appointment, color = as.factor(Treatment_Group), linetype = Pre_Post)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  labs(
    title = "Pre and Post Trends for Respondent Having Made a Primary Care Appointment in 
  Past Year",
    x = "Year",
    y = "Response to Making a PCP Appointment (1 = Yes, 0 = No)",
    color = "Treatment Group",
    linetype = "Pre/Post"
  ) + 
  scale_color_manual(values = c("mediumblue", "deepskyblue1")) +
  scale_color_manual(
    values = c("mediumblue", "deepskyblue1"),
    labels = c("Non-Medicaid", "Medicaid")  # Labeling groups manually
  ) +
geom_vline(xintercept = 2014, linetype = "dotted", color = "black", size = 0.5) +
  geom_text(aes(x = 2014, y = -Inf, label = "Medicaid Expansion"), 
            angle = 0, hjust = 0.46, vjust = -0.5, size = 3, color = "firebrick2") +
  theme_minimal()
```


**Regressions**


```{r}
# Naive Doctor Visits

did_model_doctor_visits = lm(Number_Doctor_Visits ~ Treatment_Group * Pre_Post, data = meps_data)
summary(did_model_doctor_visits)
```

```{r}
# Naive Have Usual Source of Care
did_model_haveus = lm(Have_Usual_Source_of_Care ~ Treatment_Group * Pre_Post, data = meps_data)
summary(did_model_haveus)
```

```{r}
# Naive Appointments
did_model_appointment = lm(Made_Appointment_for_Routine_Care ~ Treatment_Group * Pre_Post, data = meps_data)
summary(did_model_appointment)
```

```{r}
# Robust Number of Doctor Visits

robust_did_model_doctor_visits = lm(Number_Doctor_Visits ~ Treatment_Group * Pre_Post + relevel(factor(Region), ref = "South") + Age + Income + relevel(factor(Race), ref = "White") + relevel(factor(Sex), ref = "Male") + Years_of_Education + Employment_Status, data = meps_data)

summary(robust_did_model_doctor_visits)

```


```{r}
# Robust Have Usual Source of Care

robust_did_model_haveus = lm(Have_Usual_Source_of_Care ~ Treatment_Group * Pre_Post + Treatment_Group * Pre_Post + relevel(factor(Region), ref = "South") + Age + Income + relevel(factor(Race), ref = "White") + relevel(factor(Sex), ref = "Male") + Years_of_Education + Employment_Status, data = meps_data)

summary(robust_did_model_haveus)

```


```{r}
# Robust Appointment

robust_did_model_appointment = lm(Made_Appointment_for_Routine_Care ~ Treatment_Group * Pre_Post + Treatment_Group * Pre_Post + relevel(factor(Region), ref = "South") + Age + Income + relevel(factor(Race), ref = "White") + relevel(factor(Sex), ref = "Male") + Years_of_Education + Employment_Status, data = meps_data)

summary(robust_did_model_appointment)
```

```{r}
# Create Regression Output Table



```

```{r}
stargazer(did_model_doctor_visits, type = "text")

stargazer(robust_did_model_doctor_visits, type = "text")

```

